---
title: "An Analysis of the Australian Road Deaths Database (1989 to 2018)"
author: "Joshua Myers"
date: 
output: html_document
fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract
asdf

# Introduction
Road deaths bla bla...
Aim is to investigate road fatality data in Australia - trends over time, state, age, gender road user. 
This report uses the road fatality data (ref), supplemented with data from A, b, c (state pop, road user, etc)
Need to say it is structured data set

# Data
The Australian Road Deaths Database (ARDD) is published each month by the Bureau of Infrastructure, Transport and Regional Economics. This report uses the "Fatalities" dataset, which is a record of the number of fatalities resulting from vehicle crashes reported by the police to the State and Territory road safety authorities. This report uses the "ARDD Fatalities April 2019" data, which contains data from 50,160 road deaths that occured on Australian roads between January 1989 to May 2019. Each observation represents a death and includes a crash ID, state, month, year, day of the week, time, crash type (e.g., single, multiple, pedestrian), whether a bus, heavy rigid truck, or articulated truck was involved in the fatality, the speed limit, the type of road user killed (e.g., driver, passenger, pedestrian), gender, age, three variables relating to where the accident occured (e.g., whether it was a remote area, the name of the area), the type of road (e.g., local road or highway), and whether the accident occured during the Easter or Christmas period.

I used some other data sources to supplement the ARDD dataset for this analysis. State population over time data from the Australian Bureau of Statistics Australian Historical Population Statistics, 2019, Table 1.3: Population and components of change, for the states and territories, form 1971 to 2016 which was downloaded as an Excel file, and imported using the "read_excel" function from the *readxl* package. Population projections for 2017 and 2018 were obtained  for 2017, 2018 , Population Projections, Australia, 2017 (base) – 2066 downloaded from (https://www.abs.gov.au/AUSSTATS/abs@.nsf/Lookup/3222.0Main+Features12017%20(base)%20-%202066?OpenDocument) again, as an Excel file, and imported using read_excel. I extrated the population data for each state and the total population from 1989 to 2016 from the historical Australian Historical Population dataset, and added to this the projected populations for each state and the nation from the Population Projections data. The dataset with the state popualation data from 1989 to 2018 is referred to as the "population dataset" in this report. 

The numbers of vehicle types in the general population was obtained from the Australian Bureau of Statistics Motorvehicle Census, 2018 (https://www.abs.gov.au/AUSSTATS/abs@.nsf/DetailsPage/9309.031%20Jan%202018?OpenDocument), again downloaded as Excel file. Data from Table 1 was used for 2018, which reports the number of various different types of vehicles for the 19.2 million registered motor vehicles in Australia. This dataset is referred to as the "motorvehicle dataset" in this report. 

proportion of population each age in 2016: https://profile.id.com.au/australia/five-year-age-groups - This is referred to as the "age population" dataset in this report. 
males are approximately 50% of population in this age group https://www.abs.gov.au/ausstats/abs@.nsf/0/1CD2B1952AFC5E7ACA257298000F2E76?OpenDocument. 

These population-level data were used with the ARDD dataset to create proportions for various categories, in order to identify groups with higher risk of death, for example, by comparing the proportion of young men aged 20 to 39 who died in motorvehicle accidents in a given year with the proportion of young men that age in the general population. Exactly how these additional datasets were used in conjunction with the ARDD data is described in the Methods section. 

```{r, echo=F}
library(tidyverse)
library(ggthemes)
library(lubridate)
library(mice)
library(cowplot)
library(readxl)
theme_set(theme_minimal())

# Import road fatalities data and impute missing
# from here: https://www.bitre.gov.au/statistics/safety/fatal_road_crash_database.aspx
path = 'https://data.gov.au/data/dataset/5b530fb8-526e-4fbf-b0f6-aa24e84e4277/resource/fd646fdc-7788-4bea-a736-e4aeb0dd09a8/download/bitre_ardd_fatalities_apr_2019.csv'
road_data = read.csv(path, header = T, na.strings = c("", -9, 'Other/-9'))
# filter out 2019 data - not a complete year
road_data = filter(road_data, Year != 2019)
# Don't need Crash ID
road_data = road_data %>% select(-Crash.ID)
road_data$State = factor(road_data$State, levels=c('NSW', 'Vic', 'Qld', 'WA', 'SA', 'Tas', 'ACT', 'NT'))
road_data$Month = as.integer(road_data$Month)
road_data$Dayweek = factor(road_data$Dayweek, levels = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday'))
road_data$Bus.Involvement = as.factor(road_data$Bus.Involvement)
road_data$Heavy.Rigid.Truck.Involvement = as.factor(road_data$Heavy.Rigid.Truck.Involvement)
road_data$Articulated.Truck.Involvement = as.factor(road_data$Articulated.Truck.Involvement)
road_data$Gender[road_data$Gender == 'Unspecified'] = NA
road_data$Gender = droplevels(road_data$Gender)
road_data$Gender = as.factor(road_data$Gender)
road_data$Road.User = as.factor(road_data$Road.User)
#describe(road_data)
# state 0 missing
# month 0 missing
# year 0 missing
# day 0 missing
# time 40 missing 
# crash type 0 missing 
# bus involvement 0 missing
# heavy rigid truck involvement many missing (20493) - but I only use 2018 data
truck_2018 = road_data %>% 
  filter(Year==2018) %>% 
  summarise(count = sum(is.na(Heavy.Rigid.Truck.Involvement)))
#truck_2018 # none missing for 2018
# articultated truck involevelment 0 missing
# speed limit 1340 missing
# road user 77 missing
# gender 22 missing
# age 84 missing
# national remoteness area 45567 missing - drop this column
# national LGA Name 2017 45553 missing - drop - not using
# national road type - 45564 missing - drop
# Xmas period 0 missing
# Easter period 0 missing

# how many observations with missing data - gender, time, road user or age
missing_observations = road_data %>% 
  filter(is.na(Gender) | is.na(Time) | is.na(Road.User) | is.na(Age)) %>% 
  select(State:Time, Bus.Involvement:Articulated.Truck.Involvement, Road.User:Age)

# create time_of_day
time_of_day <- hm(road_data$Time, quiet = T)
road_data$time_of_day <- hour(time_of_day) 

road_data = road_data %>% 
  select(c(State:Dayweek, Bus.Involvement:Articulated.Truck.Involvement, Road.User:Age, time_of_day))
road_data = rename(road_data, bus_involve=Bus.Involvement, heavy_rigid_involve=Heavy.Rigid.Truck.Involvement, 
                          articulated_involve=Articulated.Truck.Involvement, road_user=Road.User) 
```

# Methods
This analysis was done using R version x (cite), in RStudio (version y), supplemented by the *tidyverse*, *ggthemes*, *mice* and *lubridate* packages (ref). The *tidyverse* is a metapackage consisting of x,yz packages. 

The dataset was downloaded as a CSV file from (put this as a ref) https://data.gov.au/data/dataset/5b530fb8-526e-4fbf-b0f6-aa24e84e4277/resource/fd646fdc-7788-4bea-a736-e4aeb0dd09a8/download/bitre_ardd_fatalities_apr_2019.csv

Opened using the 'read.csv' function in base R (cite - with version). Missing data was coded as '-9' or 'Other/-9' in the dataset. I also set Gender values of 'Unspecified' as missing (NA), since there was only one of these.  
This analysis uses data from Jan 1989 to December 2018. Data from 2019 was removed because it is not yet complete, and much of the present analysis counts data for each year. filtered out using the filter function in the dplyr package. 

This analysis used the variables: State, Month, Year, Day of the Week, Time of Day, Bus Involvement, Heavy Rigid Truck Involvement, Articultaed Truck Involvement, Road User, Gender, and Age. Other variables were removed from the dataset using dplyr's 'select' function. The variable State was coded as an orded factor with levels set to decreasing State population size using the 'factor' function. Month was set to be an ordered factor, integer using 'factor' in base R, and Day of the Week was also coded as an ordered factor from Monday to Sunday. Gender, Bus Involvement, Heavy Rigid Truck Involvement and Articulated Truck involvement were coded as binary factors using the 'as.factor' funtion in base R. Road User was coded as an unordered factor using 'as.factor',  The variable Time was converted to a numeric variable indicating the hour of the day in 24-hour time that the accident happened. The Time variable was first set to hour and minite format using the 'hm' function from the *lubridate* package, and then the hour was extracted using the 'hour' function. 

```{r, echo=F}
# Impute missing 
init = mice(road_data, maxit=0) # need to impute road user (77), gender (22), age (84), time (40)
meth = init$method
predM = init$predictorMatrix

# don't use bus involve, heavy rigid involve, or articulated involve pas predictor variables
predM[, c('bus_involve')]=0
predM[, c('heavy_rigid_involve')]=0
predM[, c('articulated_involve')]=0

# don't need to impute heavy rigid
meth[c('heavy_rigid_involve')]=""

# set methods for imputation
meth[c('road_user')]="polyreg" 
meth[c('Gender')]="logreg" 
meth[c('Age')]="pmm"
meth[c('time_of_day')]="pmm"

imputed = mice(road_data, method=meth, predictorMatrix=predM, m=5)
#densityplot(imputed) # can I somehow plot the categorical variables? see https://web.maths.unsw.edu.au/~dwarton/missingDataLab.html
#plot(imputed)
road_data <- complete(imputed)
#sapply(road_data, function(x) sum(is.na(x)))
```

```{r}
dim(road_data)
```
State, Month, Year, Day of the Week, Bus Involvement, Articulated Truck Involvement had no missing values. Time of day had 40 missing, Road User 77 missing, Gender 22 missing, and age 84 missing. In total, there were 218 observations with missing data for one of these variables. imputed using multiple imputation using the 'mice' function from the *mice* package. Missing values for Age and Hour of the Day were imputed using predictive mean matching, Gender with logistic regression imputation, and Road User with polytomous regression imputation. Diagnositic plots of imputed values looked satisfactory, and are included in Appendix A. Heavy Rigid Truck had 20,493 missing values, but I only use data from 2018 for this variable in this analysis, of which none were missing, so this variable was left as is. The imputed dataset had 11 variables providing information on 49,730 motorvehicle deaths between January 1989 and December 2018. I refer to this dataset as "the dataset" in this report. 

```{r figs, echo=F, fig.width=7, fig.height=6, fig.cap="\\label{fig:figs}Plots exploring the distributions of the variables Year, Month, Day, Hour, Age, Gender, State, and Road User "}
# summary statistics for bus, heavy_rigid and articulated involve
b = summary(road_data$bus_involve)
r = summary(road_data$heavy_rigid_involve)
a = summary(road_data$articulated_involve)

# plots showing the variation in Age, Gender, State, Year, Month, Dayweek, time_of_day, road_user, bus_involve, heavy_rigid_involve, 
age_dist = ggplot(road_data, aes(Age)) +
  geom_density() +
  ylab('Density')

sex_p = ggplot(road_data, aes(Gender)) +
  geom_bar() +
  ylab('Count') 

state_p = ggplot(road_data, aes(State)) +
  geom_bar() + # interesting NT has more than act or tas even though lowest population (ordered by population)
  ylab('Count') 

year_p = ggplot(road_data, aes(Year)) +
  geom_density() +
  ylab('Density')

month_p = ggplot(road_data, aes(as.factor(Month))) +
  geom_bar() +
  ylab('Count') +
  xlab('Month of the Year') 

day_p = ggplot(road_data, aes(Dayweek)) +
  geom_bar() +
  scale_x_discrete(labels=c("M", "Tu", "W", "Th", "F", "Sa", "Su")) +
  ylab("Count") +
  xlab("Day of the Week") 

time_p = ggplot(road_data, aes(time_of_day)) +
  geom_density() +
  ylab("Density") +
  xlab("Hour of the Day (24 hour)")

road_data$road_user = fct_infreq(road_data$road_user) 
user_p =  ggplot(road_data, aes(road_user)) +
  geom_bar() +
  ylab("Count") +
  xlab("Type of Road User")
  
plot_grid(year_p, month_p, day_p, time_p, age_dist, sex_p, state_p, user_p, labels=c(LETTERS[1:8]))
```

Figure 1 shows exploratory plots for the variables Year, Month, Day of the Week, Hour of the Day, Age, Gender, State, and Type of Road User. Numeric variables are plotted as a distribution, and factors as bar plots, which count the number of occurences for each factor level. There are no apparent issues with the data (e.g., factors have the expected categories), or obviously incorrect outliers, for example, the maximum year is 2018, and maximum hour is 23, as expected. The plot for Year shows that the number of deaths has significantly decreased over time. Friday to Sunday have more deaths than other days of the week, and the afternoon appears to be the most dangerous time to be on the road, with a secondary peak in the distribution at around midnight. Young people seem to be at much more at risk of dying, with a large peak in the distribution for age at around 20 years. Over twice as many males dying than females. The plot for state generally shows that the number of deaths decreases as state population decreases (the x-axis is ordered by population size largest to smallest from left to right), with the exception of the Northern Territory (NT), which has the smallest population, but more deaths than the Australian Capital Territory (ACT) and Tasmania. Buses were involved in 941 deaths, and articulated trucks were involved in 5071 deaths. Heavy rigid trucks were involved in 1400 deaths, although, as mentioned, there were many missing values for this variable (>20,000). 

```{r, echo=F}
# make an age discrete variable
# 0-17, 18-40, 40-60, 60+
road_data$Gender = factor(road_data$Gender, levels = c('Male', 'Female'))
road_data = road_data %>% 
  mutate(age_cat = cut(Age, breaks = c(-Inf, 16, 40, 60, Inf), labels = c('0-16 years', '17-40 years', '41-60 years', '>60 years'))) 

# create 5-year intervals as factor for grouping
five_years = c('1989 to 1993', '1994 to 1998', '1999 to 2003', '2004 to 2008', '2009 to 2013', '2014 to 2018')
road_data = road_data %>% 
  mutate(five_yr_interval = cut(Year, breaks = c(-Inf, 1993, 1998, 2003, 2008, 2013, Inf),
                                labels=five_years)) 
```

To prepare the data for analysis, I created two new discrete variables from the continuous variables Age and Year. I made a variable Age Group by splitting Age into four groups, 0 to 16, 17 to 40, 41 to 60 and >60 years, and another 5-Year Interval by dividing Year into five-year intervals. These variables were created using the "mutate" function from the *dplyr* package, in conjunction with the "cut" function from base R. The new discrete variables were used in some of the plots, for example to show the effect of age on the number of deaths over time for different age groups. 

```{r, echo=F}
# Import state population data
# from here: https://www.abs.gov.au/AUSSTATS/abs@.nsf/DetailsPage/3105.0.65.0012016?OpenDocument
state_pop = read_excel('data/historical_state_pop.xls', sheet=4, skip = 4)
# the order is nsw, vic, qld, sa, wa, tas, nt, act  
names(state_pop)[1] = 'States'
state_pop = state_pop %>% 
  filter(str_detect(States, pattern="Estimated"))
States = c('NSW', 'Vic', 'Qld', 'SA', 'WA', 'Tas', 'NT', 'ACT', 'Total')
state_pop$States = States
# only need from 1989
state_pop = select(state_pop, c(1, 20:47))
# need to get pop for 2017, 2018
# from pop projections https://www.abs.gov.au/AUSSTATS/abs@.nsf/Lookup/3222.0Main+Features12017%20(base)%20-%202066?OpenDocument
# used Series B projections, which is based on current birth rate
nsw = c(7867052,	8001204)
vic = c(6320292,	6465890)
qld = c(4928412,	5013437)
sa = c(1723714,	1735172)
wa = c(2574762,	2598092)
tas = c(522279,	526930)
nt = c(247659,	249796)
act = c(411986,	420667)
total = c(24600777,	25015825)
state_pop_17_18 = rbind.data.frame(nsw, vic, qld, sa, wa, tas, nt, act, total)
names(state_pop_17_18) = c('2017', '2018')
state_pop = cbind(state_pop, state_pop_17_18)

# get proportion of vehicles types in general population
# from 2018 mv census https://www.abs.gov.au/AUSSTATS/abs@.nsf/Lookup/9309.0Main+Features131%20Jan%202018?OpenDocument
mv_names = c('Passenger', 'CamperVan', 'LightCommercial', 'LightRigid', 'HeavyRigid', 'Articulated', 'Non-Freight', 'Bus', 'Motorcycle', 'Total')
mv_census = c(14330432, 66594, 3187131, 158032, 346966, 100694, 24165, 98565, 860700,  19173279)    
# levels of road user are: Car/Truck, Cyclist, Motorcycle, Pedestrian - so compine into car/truck/bus vs MC
# are MC more at risk - can then normalize - remove cyclist, pedestrian etc from User
names(mv_census) = mv_names

mc = mv_census[9]
total = mv_census[10]
mc_prop = mc/total # so MC account for 5% of vehicles on the road
bus = mv_census[8]
bus_prop = bus/total
heavy_rigid = mv_census[5]
heavy_rigid_prop = heavy_rigid/total
articulated = mv_census[6]
articulated_prop = articulated/total
prop_pop = rbind(mc_prop, bus_prop, heavy_rigid_prop, articulated_prop)
```

For the analysis, I first grouped the dataset by Year, State, Age Group, and Gender, using dplyr's "group_by" function, counted the numbers in each plot using dplyr's "summarise" function in conjunction with the "n" helper function, then plotted the number of deaths as a function of Year, faceted by Age and Gender using the ggplot function from the ggplot2 package (ref). I then created a dataset that counted the number of deaths for each state, and also the total number of deaths, using the group_by and tally functions from dplyr. Joined this dataset to the population dataset using dplyr's "full_join" function, after reshaping the population dataset into long form using tidyr's "gather" function. I used this dataset to create a variable, the "Number of Deaths per 100,000 people" for each state, by dividing the number of deaths each for each state by the population for that year, then multiplying by 100,000, using dplyr's mutate function. This data was then plotted using ggplot2. I then plotted the number of deaths for each age group as a function of Year for males and females, and used the Age 

# Results and Discussion
asdf

```{r, echo=F}
# Fig 1
# year, month, day by state plots
state_year_p = road_data %>% 
  #mutate_at(vars(Year, State), factor) %>%
  group_by(Year, State, age_cat, Gender) %>% 
  summarise(n=n()) %>%
  ggplot(aes(x=Year, y=n, group=State, colour=State, fill=State)) + 
  geom_line(alpha=0.5) +
  geom_smooth(se=F, size=0.5, method = 'loess') +
  ylab('Number of Deaths') +
  scale_y_continuous(trans = 'log10') +
  scale_x_continuous(breaks = c(1990, 1995, 2000, 2005, 2010, 2015), labels = c('1990', '1995', '2000', '2005', '2010', '2015')) +
  scale_colour_colorblind() +
  scale_fill_colorblind() +
  ggtitle('Number of Deaths by State (1989 to 2018)') +
  facet_grid(age_cat~Gender) +
  theme(plot.title = element_text(face='bold', hjust = 0.5, vjust = 0.5)) 
state_year_p
```

```{r, echo=F}
# Make df deaths grouped by State
deaths_by_year_state = road_data %>% 
  group_by(State, Year, age_cat, Gender) %>% 
  tally(name = 'Deaths') 

total_deaths = road_data %>% 
  group_by(Year, age_cat, Gender) %>% 
  tally(name = 'Deaths') 

total_deaths$State = 'Total'
deaths_by_year_state = rbind.data.frame(deaths_by_year_state, total_deaths)
deaths_by_year_state$State = factor(deaths_by_year_state$State, levels = c('NSW', 'Vic', 'Qld', 'WA', 'SA', 'Tas', 'ACT', 'NT', 'Total'))

state_pop_long = state_pop %>% 
  gather(key='Year', value='Population', 2:31)
names(state_pop_long)[1] = 'State'
state_pop_long$Year = as.numeric(state_pop_long$Year)
state_pop_long$State = factor(state_pop_long$State, levels = c('NSW', 'Vic', 'Qld', 'WA', 'SA', 'Tas', 'ACT', 'NT', 'Total'))

# join deaths with state population info (so can normalize)
death_pop = full_join(deaths_by_year_state, state_pop_long, by=c('State', 'Year'))

# remove total for these graphs
death_pop_states = death_pop %>% 
  filter(State !='Total') 
death_pop_states$State = fct_drop(death_pop_states$State)

# Normalize for state population 
death_pop_states = death_pop_states %>% 
  mutate(death_proportion = Deaths / Population) %>% 
  mutate(death_prop_per_100k = death_proportion * 100000)

# figure 2 deaths for each state per 100k people
prop_p_100k_sex_age = ggplot(death_pop_states, aes(x=Year, y=death_prop_per_100k, colour=State)) + 
  geom_line(alpha=0.5) +
  geom_smooth(method = 'loess', se=F, size=0.5) +
  ylab('Number of Deaths per 100,000') +
  scale_x_continuous(breaks = c(1990, 1995, 2000, 2005, 2010, 2015), labels = c('1990', '1995', '2000', '2005', '2010', '2015')) +
  scale_colour_colorblind() +
  ggtitle('Number of Deaths per 100,000 people (1989 to 2018)') +
  theme(plot.title = element_text(face='bold', hjust = 0.5, vjust = 0.5)) +
  scale_y_log10() +
  facet_grid(age_cat~Gender)
prop_p_100k_sex_age # NT young people especially males are overrepreseted
```

```{r, echo=F}
# Fig 3, age plot
age_p = ggplot(road_data, aes(x=Year, colour=age_cat, linetype=Gender)) +
  geom_line(stat = 'count') +
  scale_colour_colorblind() +
  scale_x_continuous(breaks = c(1990, 1995, 2000, 2005, 2010, 2015), labels = c('1990', '1995', '2000', '2005', '2010', '2015')) +
  labs(colour='Age Group') +
  ylab('Number of Deaths') +
  ggtitle('Number of Deaths Each by Age and Gender') +
  theme(plot.title = element_text(face='bold', hjust = 0.5, vjust = 0.5)) 
age_p # young men dominate the numbers - seem overrepresented - check against population stats for each age group - improving, but still highest 
```

```{r, echo=F}
# proportion of population each age in 2016: https://profile.id.com.au/australia/five-year-age-groups
zero_to_4 = 0.06
five_to_9 = 0.06
ten_to_14 = 0.06
fifteen_to_19 = 0.06
twenty_to_24 = 0.07
twenty5_to_29 = 0.07
thirty_to_34 = 0.07
thirty5_to_39 = 0.07
forty_to_44 = 0.07
forty5_to_49 = 0.07
fifty_to_54 = 0.07
fifty5_to_59 = 0.06
sixty_to_64 = 0.06
sixty5_to_69 = 0.05
seventy_to_74 = 0.04
seventy5_to79 = 0.03
eighty_to_84 = 0.02
eighty5_and_over = 0.02

# so 15 to 39 
prop_twenty_to_39_male = (twenty_to_24 + twenty5_to_29 + thirty_to_34 + thirty5_to_39)/2
# 14% of population 20 to 39 male in 2016
# (males are approximately 50% of population in this age group)
# https://www.abs.gov.au/ausstats/abs@.nsf/0/1CD2B1952AFC5E7ACA257298000F2E76?OpenDocument
# but accounted for what proportion of deaths?
twenty_to_39_deaths = road_data %>% 
  filter(Year==2016) %>% 
  group_by(Age, Gender) %>% 
  tally() 

total_deaths_2016 = twenty_to_39_deaths %>% 
  summarise(total=sum(n)) %>% 
  summarise(total=sum(total)) # 1293 deaths in 2016

age_names = c('0 to 19', '20 to 39', '>=40' )
age_deaths_2016 = twenty_to_39_deaths %>% 
  mutate(Age_20_39 = cut(Age, breaks = c(-Inf, 19, 39, Inf), labels = age_names)) %>% 
  group_by(Age_20_39, Gender) %>% 
  tally()
twenty_to_39_male_deaths_2016 = filter(age_deaths_2016, Age_20_39=='20 to 39', Gender=='Male') 
# 364 deaths male (110 female)
Prop_male_deaths_2016 = twenty_to_39_male_deaths_2016$n/total_deaths_2016$total # 28% of deaths, but only 14% of population
```



```{r}
# Fig 4.
# month~age~gender faceted by 1/2 decades
mth_p = road_data %>%
  group_by(Month, five_yr_interval, age_cat, Gender) %>%
  summarise(count=n()) %>%
  mutate(five_yr_ave = count/5)  %>%
  ggplot(aes(x=Month, y=five_yr_ave, group=five_yr_interval, colour=five_yr_interval)) +
  geom_line() +
  ylab('Number of Deaths') +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_log10() +
  scale_colour_colorblind() +
  scale_fill_colorblind() +
  labs(fill="Five-Year Interval") +
  labs(colour='Five-Year Interval') +
  facet_grid(age_cat~Gender) +
  ggtitle('Number of Deaths by Month') +
  theme(plot.title = element_text(face='bold', hjust = 0.5, vjust = 0.5)) 
plot(mth_p) # trend upwards in December
```


```{r, echo=F}
# fig 5.
# day~age~gender of week
day_p = road_data %>% 
  group_by(Dayweek, five_yr_interval, age_cat, Gender) %>% 
  summarise(count=n()) %>% 
  mutate(five_yr_ave = count/5) %>% 
  ggplot(aes(x=Dayweek, y=five_yr_ave,  group=five_yr_interval, colour=five_yr_interval)) +
  geom_line() +
  ylab('Number of Deaths') +
  xlab('Day of Week') +
  scale_x_discrete(labels = c('Mon', 'Tue', 'Wed', 'Thu', 'Fri', 'Sat', 'Sun')) +
  labs(colour="Five-Year Interval") +
  scale_colour_colorblind() +
  facet_grid(age_cat~Gender) +
  ggtitle('Number of Deaths by Day of the Week') +
  theme(plot.title = element_text(face='bold', hjust = 0.5, vjust = 0.5)) 
plot(day_p) # the peak of weekend deaths is dominated by the young men

```

```{r, echo=F}
# Fig 6
road_data$hour_int = as.integer(road_data$time_of_day)  
# 12 pm to 11am for plotting
road_data$hour_int = factor(road_data$hour_int, levels = c(0:23)) 
summary(road_data$hour_int)

hour_p = road_data %>% 
  group_by(hour_int, five_yr_interval, age_cat, Gender) %>% 
  summarise(count=n()) %>% 
  ggplot(aes(x=hour_int, y=count, group=five_yr_interval, colour=five_yr_interval)) +
  geom_line(alpha=0.5) +
  scale_x_discrete(breaks = c(0,3,6,9,12,15,18,21), labels = c('0', '3', '6', '9', '12', '15', '18', '21')) +
  xlab('Time (24 hour)') +
  geom_smooth(se=F, size=0.5, method = 'loess') +
  ylab('Number of Deaths') +
  scale_color_colorblind() +
  facet_grid(rows=vars(age_cat), cols = vars(Gender)) +
  ggtitle('Number of Deaths by Hour of the Day') +
  theme(plot.title = element_text(face='bold', hjust = 0.5, vjust = 0.5)) 
plot(hour_p)  # big peaks ~midnight for males 17-40, lesser so for females
```

```{r, echo=F}
# fig 7
hour_day_p = road_data %>% 
  group_by(hour_int, Dayweek, age_cat, Gender) %>% 
  summarise(count=n()) %>% 
  ggplot(aes(x=hour_int, y=count, group=age_cat, colour=age_cat)) +
  geom_line(alpha=0.5) +
  scale_x_discrete(breaks = c(0,6,12,18), labels = c('0', '6', '12', '18')) +
  xlab('Time (24 hour)') +
  geom_smooth(se=F, size=0.5, method = 'loess') +
  labs(colour='Age') +
  ylab('Number of Deaths') +
  scale_color_colorblind() +
  facet_grid(rows=vars(Gender), cols = vars(Dayweek)) +
  ggtitle('Number of Deaths by Day and Time') +
  theme(plot.title = element_text(face='bold', hjust = 0.5, vjust = 0.5)) 
plot(hour_day_p)  # big peaks ~midnight for males 17-40, lesser so for females
```


```{r, echo=F}
# road user
#summary(road_data$road_user)
road_data = road_data %>% 
  mutate(User = ifelse(road_user=='Motorcycle pillion passenger'|road_user=='Motorcycle rider', 'Motorcycle',
                       ifelse(road_user=='Driver'|road_user=='Passenger', 'Car/Truck', 
                              ifelse(road_user=='Pedestrian', 'Pedestrian', 
                                     ifelse(road_user=='Pedal cyclist', 'Cyclist', NA))))) 
road_data$User = factor(road_data$User, levels = c('Car/Truck', 'Motorcycle', 'Pedestrian', 'Cyclist'))
# fig 8
user_p = road_data %>% 
  group_by(Year, User, age_cat, Gender) %>% 
  summarise(n=n()) %>%
  ggplot(aes(x=Year, y=n, colour=User)) +
  geom_line(alpha=0.5) +
  geom_smooth(se=F, size=0.5, method = 'loess') +
  scale_x_continuous(breaks = c(1990, 1995, 2000, 2005, 2010, 2015), labels = c('1990', '1995', '2000', '2005', '2010', '2015')) +
  scale_y_log10() +
  ylab('Number of Deaths') +
  scale_color_colorblind() +
  facet_grid(age_cat~Gender) +
  ggtitle('Number of Deaths by Type of Road User') +
  theme(plot.title = element_text(face='bold', hjust = 0.5, vjust = 0.5)) 
plot(user_p)
```

```{r, echo=F}
# what proportion of deaths in 2018 were MC?
# remove pedestrian and cyclist
data_2018 = road_data %>% 
  filter(Year==2018, User != 'Pedestrian', User != 'Cyclist') 

mc_2018 = summary(as.factor(data_2018$User))
car_truck_2018 = mc_2018[1]
mc_2018 = mc_2018[2]
total_2018 = car_truck_2018 + mc_2018

prop_car_truck_deaths_2018 = car_truck_2018/total_2018
mc_prop_2018 = mc_2018/total_2018 # Only 5% of vehicles on the road are mc, but they account for 20% of mv deaths 

# proportion of trucks involved in crash compared with census 2018
# bus, heavy rigid, articulated trucks
# are any of these overrepresented?
df_2018 = road_data %>% 
  filter(Year==2018) 
bus_deaths_2018 = summary(as.factor(df_2018$bus_involve))[2]
total = nrow(df_2018)
bus_prop_2018 = bus_deaths_2018/total # 0.02 of deaths, but 0.005 of vehicles on road # not heavily overrepresented (4 times as many) - and driving a lot of the time

heavy_rigid_deaths_2018 = summary(as.factor(df_2018$heavy_rigid_involve))[2]
heavy_rigid_prop_2018 = heavy_rigid_deaths_2018 / total # 0.06 of deaths (6%) and account for 0.018 (~2%) of vehicles (overrepresented 3 times as many)

articulated_deaths_2018 = summary(as.factor(df_2018$articulated_involve))[2]
articulated_prop_2018 = articulated_deaths_2018 / total # 0.079 of deaths (~8%) and account for 0.005 of vehicles (heavily overrepresented 16 times as many)

deaths_2018 = c(mc_prop_2018, bus_prop_2018, heavy_rigid_prop_2018, articulated_prop_2018)
props_2018 = cbind.data.frame(prop_pop, deaths_2018)
names(props_2018) = c('Population', 'Deaths')
props_2018$user = c('Motorcycle', 'Bus', 'Heavy Rigid Truck', 'Articulated Truck')

# fig 9
# need to make long
props_2018_p = props_2018 %>% 
  gather('Group', 'Proportion', 1:2) %>% 
  ggplot(aes(x=factor(user, levels=c('Bus', 'Heavy Rigid Truck', 'Articulated Truck', 'Motorcycle')), y=Proportion, colour=Group, fill=Group)) +
  geom_bar(stat='identity', position = 'dodge') +
  scale_color_colorblind() +
  scale_fill_colorblind() +
  ggtitle('Proportion of Vehicle Types Involved in Fatatities \n Compared with Overall Population') +
  xlab('Type of Vehicle') +
  theme(plot.title = element_text(face='bold', hjust = 0.5, vjust = 0.5)) 
plot(props_2018_p)
```

# Conclusions
asdf

# References

# Appendix A: MICE Plots

# Appendix B: R Code



